Against Spurious Understanding: Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-01-31 main.tex
updates from OL

%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

@@

\twocolumn[
\icmltitle{Spurious Sentience:\\
           Dovelating Inflated AI World Model Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}

@@

\begin{abstract}
This document provides a basic paper template and submission guidelines.
Abstracts must be a single paragraph, ideally between 4--6 sentences long.
Gross violations will trigger corrections at the camera-ready phase.
\end{abstract}

@@

We humans are prone to seek patterns everywhere. Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future. Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g. the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.

In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other. Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation. Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones. A major contributor is that humans struggle to conceive of actual randomness: e.g. studies show that humans are poor at detecting e.g. \cite{falk1997making}, and creating e.g. \cite{ladouceur1996erroneous} random sequences. A common issue is a lack of expectation that the patterns that hint towards a causal relationship, such as correlations, will still appear at random. This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \cite{nickerson1998confirmation}), or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.

We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems. Such that, when an AI system begins finding patterns in the corpus not originally perceived during training, that is now from a human perspective, observed as greater cognitive capabilities. This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
%Such as an AI system finding patterns in data not originally perceived during training
%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g. \cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'. When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that also aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience \footnote{https://www-scientificamerican-com.tudelft.idm.oclc.org/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/} and consciousness.

Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs. We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, within AI systems. We further display results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.

\textcolor{red}{andrew: I'm thinking we can adjust this all to aim towards more 'severe' tests or something. }

\section{Background}

\subsection{Claims of Artificial General Intelligence}

\textcolor{red}{antony: I will be adding citations and other links later, just getting text down for now.}
As previously mentioned, AGI is the current objective for which the global AI community is shooting for. As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts. This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds. True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.

Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI. This model has the ability to solve complex geometry problems, which, they state it does without the need for human demonstrations during training. Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI. As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer. With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.

However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s. Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry. They were able to achieve this success due to geometry being an axiomatically closed system. Therefore, despite DeepMinds success in creating a powerfully fast Geometry solving machine, it is still far from being an AGI.

One of the more contentious pieces of literature recently studied is the paper by \cite{gurnee2023language}. This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com} (formally Twitter) platform. Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.

They claim said models generate an internal representation of the world, with which spatiotemporal representations are created. Further to this we see claim to the generalization of tasks, in which the model can answer questions regarding spatial and temporal relationships, without the need for further training. Although these can be seen as building blocks for a more comprehensive model, it is also fair to assume that it is mere conjecture and more likely spurious correlations.

As we also consider the paper by the Microsoft Research team \cite{}, despite the illusion of a claim that GPT-4 shows sparks of AGI, they do not directly claim that GPT-4 has AGI. However they do claim that we are moving closer to this goal, which is not an outlandish claim given the advancements we have seen with AI in recent years.

Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \cite{nickerson1998confirmation}. Importantly, it need not be fully conscious. Rather, theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief. Equally importantly, is that theory suggests these behaviors may be motivated or unmotivated: one may selectively seek, attend to, or interpret evidence in favor of a hypothesis for which they may or may not have a personal interest in supporting.

Hypotheses in computer science fields related to AI are often implicit. Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis. However with regards to claims that outputs of a system hint towards emerging qualities like AGI, hypotheses necessarily become more explicit: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).

Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews. Scientists may pay little to no attention to competing hypotheses or explanations, e.g. only considering the likelihood that outputs of system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false. Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a 'weak' or 'non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}. In extreme cases, propositions may be made that cannot be falsified based on how they are formulated. If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question. Thus, one is far more likely to show evidence in favor of their beliefs by posing a weak null hypotheses. Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs. As \cite{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.

\subsection{Anthropomorphism}

Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects. These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}. Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative. Anthropomorphism was implied because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review). 
Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end may attribute human emotions, cognition, and intelligence to AI systems. As \cite{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.

Modern theory of anthropomorphism suggests there are three key components, one of which is a cognitive feature, and two of which are motivations. The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: "...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}. Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors. This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}. An illustration is the example of Tom Hank's character in the film Castaway who has conversations with a ball he calls "Wilson". Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc. with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.

The motivation as a human to be competent, so-called effectance motivation, is the most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}. When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}. Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g. an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agents behaviors. This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn. One such situation is the increasing demand for 'AI' skills among AI non-experts: e.g. job postings for individuals with skills in 'GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}. Thus, individuals in advanced economies may experience a pressure to rapidly learn how to effectively use and understand an opaque technology that appears to behave human-like in at least some ways, and by imbuing the technology with humanity may in turn ease their learning curve.

These pressures extend to experts as well, in both scholarly and commercial fields. Academics face considerable pressures to publish work in the field of AI: citation metrics and scholarly publications remain the primary metric for tenure and promotion e.g. \cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced in overall scholarly publications \footnote{ Retrieved 23/01/23 from https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021 }, and in peer-reviewed publications \cite{Maslej2023-pi}, with both more than doubling from 2010 to 2020. The adoption of techniques underlying technologies with the AI label, i.e. machine learning, has spread to fields far beyond Computer Science, e.g. Astronomy and Physics to Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  >  2009  AND  PUBYEAR  <  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}. Outside of academia, the number of jobs requiring AI expertise have increased several fold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}. Given the strong incentives to be AI-competent, it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g. \cite{salles2020anthropomorphism, van2023reclaiming}.

\section{Patterns in Latent Spaces and How to Find
Them}\label{patterns-in-latent-spaces-and-how-to-find-them}

To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples. To illustrate further that this phenomenon is neither surprising nor unique to the field of computer science, we will borrow draw on ideas from the fields of Economics and Finance in this section. We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data. The final example in Section \ref{ex-llm} is a bit more involved and closer in spirit to the experiments conducted by \citet{gurnee2023language}. As we go along, we will try to discuss both the benefits and potential pitfalls of finding patterns in latent spaces.

\subsection{Example: Random Projections of World Models}\label{example-deep-learning}

The Llama-2 model has ingested huge amounts of publicly available data from the internet including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}. It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly. At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates. The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance, before being passed through a series of layers which are gradually trained for downstream tasks.

In our first example in this section, we simulate this scenario, stopping short of training the model. In particular, we

\begin{itemize}
\tightlist
\item
  Take the
  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
  that was used in \cite{gurnee2023language}, which maps locations/areas
  to their latitude and longitude. For each place, it also indicates the
  corresponding country.
\item
  Take the subset that contains countries that are currently part of the
  top 10
  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
  world ranking}.
\item
  Assign the current rank to each country, i.e.~Argentina gets 1, France
  gets 2, \ldots{}
\item
  Transform the longitude and latitude data, respectively, as follows:
  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\). This is to
  ensure that the training data only involves a noisy version of the
  coordinates.
\item
  Next, encode all features except the FIFA world rank indicator as
  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
  of samples and \(m\) is the number of resulting features.
\item
  Additionally, add a large number of random features to \(X\) to
  simulate the fact that not all features ingested by Llama-2 are
  necessarily correlated with geographical coordinates. Let \(d\) denote
  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
  of random features.
\item
  Initialize a small neural network, let's call it a \emph{projector},
  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
  units and sigmoid activation and then from there to a
  lower-dimensional output space.
\item
  Without performing any training on the \emph{projector}, we simply
  compute a forward pass of \(X\) through the \emph{projector} and
  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
\item
  Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge
  regression:
  \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\)
  where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the
  longitude and latitude for each sample. A hold-out set is reserved for testing.
\item
  Finally, we compute the predicted coordinates for each sample in the hold-out set as
  \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on
  a world map (see Figure~\ref{fig-map}).
\item 
  While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand.
\end{itemize}

@@

While this is admittedly a somewhat contrived example, it serves to illustrate the point that if we randomly project a small number of features that are highly correlated or predictive of the target into a large enough space, we should expect to be able to recover meaningful representations of the target from the latent space. Similarly, \cite{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification. In fact, it is well-known that random projections can be used for prediction tasks \cite{dasgupta2013experiments}.

\subsection{Example: Principal Component
Analysis}\label{example-principal-component-analysis}

For our next example, we will move beyond randomness and illustrate how principal component analysis (PCA) can be used to recover meaningful and compact latent representations from data in an unsupervised manner.

The yield curve is a popular tool for investors and economists to gauge the health of the economy. It plots the yields of bonds against their maturities. The slope of the yield curve is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy. While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone. To leverage this information in downstream modelling tasks, economists regularly make use of PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data. Recent examples of such include \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing} among others.

Here we are primarily interested in understanding the nature this low-dimensional projection, so will limit our attention to the US Treasury yield curve. To this end, we have collected daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury. Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). To put this into the broader context of this, let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.

>>>>

%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
\RequirePackage[hyphens]{url}
\documentclass{article}

@@

\twocolumn[
\icmltitle{Against Spurious Understanding: 
           Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}

@@

\begin{abstract}
Humans have a tendency to place a `human' like quality in the objects around them. Giving your car a name or talking to your pet, such that you believe they understand you in a human way. This behavior is also seeing traction in Machine Learning (ML), where the idea of sentience is being placed upon Large Language Models (LLMs). In this position paper we discuss the history of and why humans invariably anthropomorphize onto various objects. How the search for Artificial General Intelligence (AGI) is pushing recent literature to over emphasis sentience. To combat this, we provide a limited study, that investigates the latent space of models to find spurious correlations. Finally we discuss how the popularity in media and the race for AGI enforces over-interpretation of consciousness in the models we use.

\end{abstract}

@@

% \textcolor{magenta}{TODOs for the team:
% \begin{itemize}
%     \item Please double check the citations. Note this style file does not retain capitalization in titles. A trick to retain capitalization is to put an extra set of brackets around the title. I did this for the Ladouceur et al. and Bubeck et al references already; please make this consistent throughout.
%     \item As another issue with the citations: note that arxiv preprints do not show very well. Best way to make them show is to wrap them as 'journal articles' as done in the Arora citation. Please straighten this out for the other articles too
% \end{itemize}
% }

In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\cite{merton1942science}. While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.

At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening. Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\cite{weissburg2024tweets}. Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible only is increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.

Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence. In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal. The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}. In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or showing performance considered `surprising' to humans.

For example, Google DeepMind claimed their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI. This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training. However, as noted by Hector Zenil \cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s. Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system. Therefore, despite DeepMind's success in creating a powerfully fast geometry solving machine, it is still far from being AGI.

Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}. Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.

When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness. AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}. As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder. In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.

Noticing these trends, we as the authors of this article are concerned. We feel that the current culture of racing towards Big Outcome Statements in industry and academic publishing too much disincentivizes efforts towards more thorough and nuanced actual problem understanding. At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there indeed is sufficient evidence for these claims. With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs). In other words, it is both very tempting to claim surprising human-like achievements of AI, and as humans we are very prone to genuinely believe this.

To strenghten our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made. We work out different experiments which may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made. Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there. Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.

%Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI. As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer. With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.

%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}. This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com/wesg52/status/1709551516577902782?s=20} (formally Twitter) platform. Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.

%This only gets amplified as the interest in Artificial General Intelligence has now merged with a push towards building general-purpose tools.

%\textcolor{magenta}{cynthia: I am now moving the texts of Andrew and Antony on AGI here; maybe some tightening can already be done in this. Unfortunately no time for thorough editing yet at this stage, but will come back to this later}

%In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g. \cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'.

%Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs. We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, within AI systems. We further display results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.

%As previously mentioned, AGI is the current objective for which the global AI community is shooting for. As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts. This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds. True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.

% Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI. This model has the ability to solve complex geometry problems, which they state it does without the need for human demonstrations during training. Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI. As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer. With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.

% However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s. Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry. They were able to achieve this success due to geometry being an axiomatically closed system. Therefore, despite DeepMinds success in creating a powerfully fast Geometry solving machine, it is still far from being an AGI.

%it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g. \cite{salles2020anthropomorphism, van2023reclaiming}.

\section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}
\textcolor{magenta}{cynthia: @Patrick this section needs a little bit of tightening too. I have tried doing this, please do double-check that I didn't over-shorten things. In any case, I will always comment out a copy of the original text close to what I now put}

In Winter 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform~\cite{gurnee2023languagev1}. In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that only can correlate and mimic existing patterns from data, but not understand this data.

While the manuscript immediately received public criticism~\cite{marcus2023muddles}, and the subsequent, current manuscript version  is more careful with regard to its claims~\cite{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings. However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation. After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on economic growth prediction.

%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}. This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform. Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.

%They claim said models generate an internal representation of the world, with which spatiotemporal representations are created. Further to this we see claim to the generalization of tasks, in which the model can answer questions regarding spatial and temporal relationships, without the need for further training. Although these can be seen as building blocks for a more comprehensive model, it is also fair to assume that it is mere conjecture and more likely spurious correlations.

%In this section, we navigate a number of simple examples to support our claim, that observing patterns in latent spaces should not be a surprising revelation.
%To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples. 
%We will draw on ideas from the fields of Economics and Finance, to illustrate further that this phenomenon is neither surprising nor unique to the field of computer science. We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data. The final example in Section \ref{ex-llm} is more complex and closer in spirit to the experiments conducted by \citet{gurnee2023language}. As we progress, we will discuss both the benefits and potential pitfalls of finding patterns in latent spaces.

@@

\subsection{Do neural networks get born with world models?}\label{example-deep-learning}
%\subsection{Example: Random Projections of World Models}\label{example-deep-learning}
\textcolor{magenta}{Patrick feel free to revert, I kept all original titles}
\cite{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places. Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}. It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly. At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates. The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance, before being passed through a series of layers which are gradually trained for downstream tasks.

In our first example in this section, we simulate this scenario, stopping short of training the model. In particular, we take the
  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
  that was used in \cite{gurnee2023languagev2}, which maps locations/areas
  to their latitude and longitude. For each place, it also indicates the
  corresponding country. From this, we take the subset that contains countries that are currently part of the
  top 10
  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
  world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France
  gets 2, \ldots{}).

  Subsequently,  to
  ensure that the training data only involves a noisy version of the
  coordinates, we transform the longitude and latitude data, respectively, as follows:
  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).

  Next, we encode all features except the FIFA world rank indicator as
  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
  of samples and \(m\) is the number of resulting features. 
  Additionally, we add a large number of random features to \(X\) to
  simulate the fact that not all features ingested by Llama-2 are
  necessarily correlated with geographical coordinates. Let \(d\) denote
  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
  of random features.

  We then initialize a small neural network, which we consider as a \emph{projector},
  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
  units and sigmoid activation, and from there to a
  lower-dimensional output space. Without performing any training on the \emph{projector}, we simply
  compute a forward pass of \(X\) through the \emph{projector} and
  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
